NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2020-11-05 Manuscript/manuscriptPGPE.tex
Update manuscript

The conformational ensembles of different lipid headgroups and their order parameters have been
discussed previously \cite{??}, but these studies have been inconclusive because order parameter
signs have been unknown and tools to interpret the conformational ensembles have not been available.
Our experimental order parameters witht the information of the sign from different lipid headgroups
combined with the literature data are shown in Fig. \ref{HGorderParameters}.
The glycerol backbone order parameters are similar for all the lipids, although they move slightly toward
positive values (closer to zero) in the order PC $<$ PE $<$ PS $<$ PG. 
%Also the headgroup order parameters of PC lipids are close PE, althought the latter gives systematically
%slightly more positive values (Fig. \ref{HGorderParameters}).
% These could be explained with slightly larger temperature in PE measurements, except for the $\alpha$-carbon with the positive sign, for which the
% more positive value is farther away from zero.
While headgroup order parameters of PC and PE lipids are rather similar,
PG and PS lipids exhibit distinct values compared with other lipids.
In PS lipids, the $\alpha$-carbon order parameter exhibits significant forking and
the $\beta$-carbon order parameter is more negative than others.
The $\alpha$-carbon order parameter of PG is similar to PE and PG, while the positive value of $\beta$-carbon
is distinct from the other lipids.
Notably, this difference was not observed in previous $^2$H NMR experiments,
because absolute value of $\beta$-carbon order parameter is similar in PG, PE and PC lipids and the order parameter
signs were not measured~\cite{wohlgemuth80,gally81,borle85}.

@@

In conclusion, the order parameter experiments suggest that the glycerol backbone conformations in all lipids
and the headgroup conformations in PC and PE lipids are relatively similar, while 
PS and PG headgropus exhibit distinct conformational sampling.
The details of sampled conformation are difficult to deduce from order parameters only,
but the distinct headgroup order parameters
of PS lipids are previously related to the more rigid structure of the headgroup~\cite{browning80,buldt81,antila19}.

%\subsection{Headgroup and glycerol backbone of POPE and POPG in MD simulations}
Headgroup and glycerol backbone order parameters of PE and PG lipids
show wide variation between different force fields
and none of the force fields reproduce all values within experimental error bars.
(Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}),
as observed previously also for PC and PS lipids \cite{botan15,antila19}.
%The Slipid simulations were able to capture the essential differences between PC and PS lipid
%headgroups \cite{antila19}, but this is not the case for PE and PG lipids.
The poor performance of headgroup order parameters in Berger model can be probably explained by
ring like structures seen in Fig. 6 in Ref. \citenum{mukhopadhyay04}, which is a typical feature
for Berger based lipid force fields containing explicit hydrogen atoms in the head group \cite{zhao08,henin09,dahlberg10}.
The poor performance of glycerol backbone of Slipids simulations is systemically observed also
for other lipids in previous studies \cite{botan15,antila19}.
\todo{Should we comment more the relative quality of different force fields and/or make the subjective force field ranking figures?
https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/8}

Without further discussion about poorly performing force fields, we focus on more detailed analysis
of CHARMM36 simulations, which captures the essential differences between
PC, PS, PG and PE headroup order parameters  (Fig.~\ref{structures}) with the exception of $\beta$-carbon order parameter of PC
%give the best overall agreement with experiments for
%the headgroup and glycerol backbone order parameters of all PC, PS, PG and PE lipids in this and our previous
%studies \cite{botan15,antila19}. Even thought many order parameters in CHARMM36 simulations are
%not within the experimental error bars, it 
which is too negative when compared with PS or PE order parameter, or with experiments \cite{botan15}.
%As already pointed out in our previous study \cite{antila19},
Characteristic dihedral conformations in PS headgroup are asymmetric conformations
preferring gauche 270$^\circ$ conformations in N-C$_\beta$-C$_\alpha$-O$_\alpha$ and
%dihedral exhibits a more asymmetric angle distribution for PS, having largest probability being
%in , than for the PC headgroup.
C$_\beta$-C$_\alpha$-O$_\alpha$-P dihedrals. 
In PG headgroup, the O$_\beta$-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral (corresponding N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral in other lipids)
is mostly in trans conformation, and C$_\beta$-C$_\alpha$-O$_\alpha$-P  has asymmetric tendency to be in
gauche 60$^\circ$ conformation. 
Main difference between PC and PE is the lower probability of trans state in C$_\beta$-C$_\alpha$-O$_\alpha$-P PC dihedral,
which could be a potential reason for the too negative $\beta$ headgroup order parameter in PC.

@@

\clearpage
\subsection{Lipid conformational ensembles in lipid bilayers with bound ions}
\begin{figure*}[bt]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithCaClPG1PC1.eps}
  \caption{\label{changesWITHCaClPG}
    Modulation of headgroup order parameters of POPC ({\it left}) and POPG ({\it right}) in POPC:POPG (1:1)
    mixture upon addition of CaCl$_2$ in 298 K temperature from experiments \cite{borle85,macdonald87} and simulations.
    The $\beta$-carbon order parameter of POPC (dashed line on top left) is not directly measured but
    calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
    The changes with respect to the systems without CaCl$_2$ are shown for other data than
    for the $\alpha$-carbon of POPG for which experimental order parameter is not available.
    %(right) The headgroup order parameters of PG from PC:PG mixtures as a function CaCl$_2$
    %concentration from experiments \cite{borle85} and CHARMM36 simulations.
  }
\end{figure*}

To evaluate the calcium binding in simulations of lipid bilayers containing PG lipids,
we calculated the changes in headgroup order parameters of POPC:POPG (1:1) and (4:1) mixtures
upon addition of CaCl$_2$, and compared these with the available experimental data \cite{borle85,macdonald87}.
The headgroup order parameters of PC lipids can be used to measure the ion binding
affinity to lipid bilayers because their magnitude is linearly proportional to the amount of bound charge in bilayer \cite{seelig87,catte16}.
This molecular electrometer concept can be used also for bilayers containing PC lipids mixed with charged lipids \cite{borle85,macdonald87,roux90,antila19}.
The headgroup order parameters can be used to evaluate MD simulations against experimental data,
because they can be directly calculated from MD simulations \cite{catte16}.

The decrease of POPC headgroup order parameters in mixtures with POPG lipids with
increasing CaCl$_2$ concentration is overestimated in Slipids and Lipid17
simulations (Figs. \ref{changesWITHCaClPG} and \ref{changesWITHCaClPG1PC4})
indicating too strong binding affinity of calcium into the bilayers as previously
observed for pure PC lipid bilayers and mixtures with PS lipids \cite{catte16,antila19}.
\todo{CHARMM results to be mentioned once we have the new simulations.}
%The decrease of $\alpha$-carbon order parameter of PC lipids in PC:PG mixtures as a function
%of calcium concentration is close to experiments CHARMM36 simulations (Fig. \ref{changesWITHCaClPG}),
%but the decrease of $\beta$-carbon order parameter seems to be overestimated.
%However, the $\beta$-carbon order parameter was not actually measured from these samples,
%but they are calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
%On the other hand, the simulations are not converged.
The calcium binding affinity to lipid bilayers with PC and PS lipids
was recently improved by applying the electronic continuum correction (ECC) to Amber Lipid14/17 force fields \cite{melcr18,melcr20}.
In this approach, the electronic polarizability is implicitly included in the classical force fields
by scaling the charges with constant factors \cite{leontyev11}. Here, we make a ECC-POPG force field
by applying the scaling factors originally used for POPS also to POPG, i.e., we multiply
charges and Lennard-Jones $\sigma$s of headgroup, glycerol backbone, and carbonyl regions with 
$f_q$=0.75 and $f_\sigma$=0.89, respectively \cite{melcr20}.
ECC-POPG model gives a weaker calcium binding affinity (Fig.\ref{CAdensPG}) and better agreement with the experimental
PC headgroup order parameter changes (Fig. \ref{changesWITHCaClPG}) for POPC:POPG mixtures than the original Lipid17 model
\todo{to be finished when we have all the data}, indicating that the ECC improves the simulation predictions of
calcium binding affinity as previously observed for PC and PS lipids \cite{melcr18,melcr20}.

Experimental data for the $\beta$-carbon order parameter of POPG
shows a rapid decrease with increasing CaCl$_2$ concentrations up to 10 mM
and more modest decrease with larger concentrations (Fig. \ref{changesWITHCaClPG}) \cite{borle85}.
This behaviour is similar to that of $\beta$-carbon order parameters of POPC,
but essentially different than observed for POPS, where $\beta$-carbon order parameters increases with
addition of calcium \cite{antila19}. Experimentally measured changes of PG $\alpha$-carbon order parameters upon addition of
calcium are not available.
%upon addition of calcium in experiments,
%exhibiting 
%The behaviour is qualitatively different than in the $\beta$-carbon of PS lipids which increase upon addition of CaCl$_2$.
Lipid17 and Slipids force fields correctly capture the PG $\beta$-carbon order parameter response to CaCl$_2$
even thought the binding affinity was too large based on the comparison of PC headgroup order parameter changes with experiments.
While applying ECC to Lipid17 improved the PC headgroup order parameter response and binding affinity,
the response of PG $\beta$-carbon order parameter to calcium is too weak in this model.
The response of PG $\alpha$-carbon order parameters to CaCl$_2$ differs between force fields,
but experimental data to evaluate these predictions is not available. 
\todo{To be finished once we have the new CHARMM simulations and conformational changes of PG analyzed.}
% The result is similar to the $\sim$200~ns simulations with PC lipids in previous work \cite{catte16}.
%However, when simulation was continued for $\mu$s, the binding affinity substantially increased
%and interpretation was that calcium overbinds to PC lipid in CHARMM36. Therefore, the
%conclusion seems to be similar here, although the new NBfix parameters may complicate
%the situation \todo{The status of NBfix parameters in these simulations should be checked.}.

>>>>

Derivation of different lipid headgroup conformational ensembles in liquid state has been
%discussed previously \cite{??}, but these studies have been
inconclusive due to lack of suitable experimental data and tools to interpret the conformational ensembles.
Our experimental order parameters, including the signs, from different lipid headgroups
combined with the literature data are shown in Fig. \ref{HGorderParameters}.
The glycerol backbone order parameters are similar for all the lipids, although they move slightly toward
positive values (closer to zero) in the order PC $<$ PE $<$ PS $<$ PG. 
%Also the headgroup order parameters of PC lipids are close PE, althought the latter gives systematically
%slightly more positive values (Fig. \ref{HGorderParameters}).
% These could be explained with slightly larger temperature in PE measurements, except for the $\alpha$-carbon with the positive sign, for which the
% more positive value is farther away from zero.
While the headgroup order parameters are similar for PC and PE lipids,
PG and PS lipids exhibit distinct values in comparison with other lipids.
In PS headgroup, the $\alpha$-carbon order parameter exhibits significant forking and
the $\beta$-carbon has more negative value than in other lipids.
In PG headgroup, the $\alpha$-carbon order parameter is similar to PE and PG,
while the positive value of $\beta$-carbon is distinct from all the other lipids.
Notably, this difference was not observed in previous $^2$H NMR experiments,
because absolute value of $\beta$-carbon order parameter is similar in PG, PE and PC lipids and the order parameter
signs were not measured~\cite{wohlgemuth80,gally81,borle85}.

@@

As in previous NMRlipids project results for PC and PS lipids \cite{botan15,antila19},
none of the MD simulation force fields correctly captures all the
headgroup and glycerol backbone order parameters of PE and PG lipids
(Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG})
that would enable a straightforward interpretation of conformational ensembles.
%show wide variation between different force fields
%and none of the force fields reproduce all values within experimental error bars.
%(Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}),
%as observed previously also for PC and PS lipids \cite{botan15,antila19}.
%The Slipid simulations were able to capture the essential differences between PC and PS lipid
%headgroups \cite{antila19}, but this is not the case for PE and PG lipids.
%Without further discussion about poorly performing force fields, we focus on more detailed analysis
%of
Nevertheless, CHARMM36 force field, which gives the results closest to the experiments for all lipids,
captures the essential differences between
PC, PS, PG and PE headroup order parameters  (Fig.~\ref{structures}) with the exception of $\beta$-carbon order parameter of PC
%give the best overall agreement with experiments for
%the headgroup and glycerol backbone order parameters of all PC, PS, PG and PE lipids in this and our previous
%studies \cite{botan15,antila19}. Even thought many order parameters in CHARMM36 simulations are
%not within the experimental error bars, it 
which is too negative when compared with other lipids or experiments \cite{botan15}.
% As already pointed out in our previous study \cite{antila19},

The conformational ensembles characterized by the distributions of heavy atom dihedral angles
show major differences between lipids only for O$_\alpha$-C$_\alpha$ and  C$_\alpha$-C$_\beta$ bonds
near the end of the headgroup (Fig. \ref{structures}).
These probably explain the slight differences in the
angle between headgroup dipole and membrane normal, which decreases 
in the order PG $>$ PE  $>$ PC  $>$ PS, although difference between PC and PE may an artefact as
the $\beta$-carbon order parameter in PC is poorly reproduced by the CHARMM36 force field.
Dihedrals of both O-P bonds can freely rotate in all lipids as all possible angles
are observed in the distributions. For PS, however, some angles are quite unlikely,
possible explaining the more rigid headgroup structures proposed for PS lipids~\cite{browning80,buldt81}.
The eclipsed (0 or 360 degrees) are not present in any other dihedrals, and anti-eclipsed (120 or 240 degrees)
are not present in C$_\alpha$-C$_\beta$ and g$_1$-g$_2$ bonds.

%Characteristic dihedral conformations in PS headgroup are asymmetric conformations
%preferring gauche 270$^\circ$ conformations in N-C$_\beta$-C$_\alpha$-O$_\alpha$ and
%%dihedral exhibits a more asymmetric angle distribution for PS, having largest probability being
%%in , than for the PC headgroup.
%C$_\beta$-C$_\alpha$-O$_\alpha$-P dihedrals. 
%In PG headgroup, the O$_\beta$-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral (corresponding N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral in other% lipids)
%is mostly in trans conformation, and C$_\beta$-C$_\alpha$-O$_\alpha$-P  has asymmetric tendency to be in
%gauche 60$^\circ$ conformation. 
%Main difference between PC and PE is the lower probability of trans state in C$_\beta$-C$_\alpha$-O$_\alpha$-P PC dihedral,
%which could be a potential reason for the too negative $\beta$ headgroup order parameter in PC.%

In conclusion, the results suggest that free rotations around O-P bonds decouple the
headgroup and glycerol backbone structure and dynamics in similarly in all lipids,
althought this rotation is slower in PS lipids where crossing certain barriers are slower.
Despite significant differences in dihedral distributions near the headgroup end,
all lipids have dihedral angles within approximately the same ranges.
This suggests that all lipids can quite freely arrange to multiple headgroup conformations
when interaction with proteins, ions or other biomolecules.
%the order parameter experiments suggest that the glycerol backbone conformations in all lipids
%and the headgroup conformations in PC and PE lipids are relatively similar, while 
%PS and PG headgropus exhibit distinct conformational sampling.
%The details of sampled conformation are difficult to deduce from order parameters only,
%but the distinct headgroup order parameters
%of PS lipids are previously related to the more rigid structure of the headgroup~\cite{browning80,buldt81,antila19}.

@@

\clearpage
\subsection{Lipid conformational ensembles in lipid bilayers with bound ions}

Headgroup order parameters of PC lipids decrease or increase upon addition of positive
or negative charge into membrane, respectively, while they remain unchanged upon addition
of neutral molecules such as cholesterol or sphingomyelin \cite{??}.
This has been explained by the tilting of the headgroup dipole due to electrostatic
interactions with bound charges, and has been used to measure the binding affinity
of ions to membranes \cite{??}.
%The headgroup order parameters of PC lipids can be used to measure the ion binding
%affinity to lipid bilayers because their magnitude is linearly proportional to the amount of bound charge in bilayer \cite{seelig87,catte16}.
%This molecular electrometer concept can be used also for bilayers containing PC lipids mixed with charged lipids \cite{borle85,macdonald87,roux90,antila19}.
%The headgroup order parameters can be used to evaluate MD simulations against experimental data,
%because they can be directly calculated from MD simulations \cite{catte16}.

%To evaluate the calcium binding in simulations of lipid bilayers containing PG lipids,
%we calculated
The changes of PC headgroup order parameters in POPC:POPG (1:1) and (4:1) mixtures
upon addition of CaCl$_2$ from simulations and experiments \cite{borle85,macdonald87}
%, and compared these with the available experimental data \cite{borle85,macdonald87}.
in Figs. \ref{changesWITHCaClPG} and \ref{changesWITHCaClPG1PC4} are in line with
conclusion from our previous studies \cite{??}:  
%The decrease of POPC headgroup order parameters in mixtures with POPG lipids with
%increasing CaCl$_2$ concentration is overestimated in Slipids and Lipid17
order parameters decrease in the response to bound cations in qualitative agreement with
experiments, but calcium binding affinity to membranes is overestimated in 
simulations except in CHARMM36 with the NBfix correction which underestimates the binding affinity.
On the other hand, implicit inclusion of electronic polarizability to Amber based lipid models
improves the ion binding affinity leading to a good agreement with experiments.
%(Figs. \ref{changesWITHCaClPG} and \ref{changesWITHCaClPG1PC4})
%
%indicating too strong binding affinity of calcium into the bilayers as previously
%observed for pure PC lipid bilayers and mixtures with PS lipids \cite{catte16,antila19}.
%
%The decrease of $\alpha$-carbon order parameter of PC lipids in PC:PG mixtures as a function
%of calcium concentration is close to experiments CHARMM36 simulations (Fig. \ref{changesWITHCaClPG}),
%but the decrease of $\beta$-carbon order parameter seems to be overestimated.
%However, the $\beta$-carbon order parameter was not actually measured from these samples,
%but they are calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
%On the other hand, the simulations are not converged.
%
%The calcium binding affinity to lipid bilayers with PC and PS lipids
%was recently improved by applying the electronic continuum correction (ECC) to Amber Lipid14/17 force fields \cite{melcr18,melcr20}.
%In this approach, the electronic polarizability is implicitly included in the classical force fields
%by scaling the charges with constant factors \cite{leontyev11}. Here, we make a ECC-POPG force field
%by applying the scaling factors originally used for POPS also to POPG, i.e., we multiply
%charges and Lennard-Jones $\sigma$s of headgroup, glycerol backbone, and carbonyl regions with 
%$f_q$=0.75 and $f_\sigma$=0.89, respectively \cite{melcr20}.
%ECC-POPG model gives a weaker calcium binding affinity (Fig.\ref{CAdensPG}) and better agreement with the experimental
%PC headgroup order parameter changes (Fig. \ref{changesWITHCaClPG}) for POPC:POPG mixtures than the original Lipid17 model
%\todo{to be finished when we have all the data}, indicating that the ECC improves the simulation predictions of
%calcium binding affinity as previously observed for PC and PS lipids \cite{melcr18,melcr20}.

\begin{figure*}[bt]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithCaClPG1PC1.eps}
  \caption{\label{changesWITHCaClPG}
    Modulation of headgroup order parameters of POPC ({\it left}) and POPG ({\it right}) in POPC:POPG (1:1)
    mixture upon addition of CaCl$_2$ in 298 K temperature from experiments \cite{borle85,macdonald87} and simulations.
    The $\beta$-carbon order parameter of POPC (dashed line on top left) is not directly measured but
    calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
    The changes with respect to the systems without CaCl$_2$ are shown for other data than
    for the $\alpha$-carbon of POPG for which experimental order parameter is not available.
    %(right) The headgroup order parameters of PG from PC:PG mixtures as a function CaCl$_2$
    %concentration from experiments \cite{borle85} and CHARMM36 simulations.
  }
\end{figure*}

Changes in lipid conformational ensembles characterized by distributions of heavy atom dihedral angles
are very small upon binding of calcium or addition of cationic surfactants to membranes (Figs. \ref{??}).
Yet, such changes are sufficient to tilt the headgroup dipole angle and reproduce the experimentally observed
order parameter changes in $\alpha$ and $\beta$ carbons.

Simulations also qualitatively reproduce the experimentally observed decrease of POPG
%Experimental data for the
$\beta$-carbon order parameter
%of POPG shows a rapid decrease with increasing
upon addition of CaCl$_2$,
%concentrations up to 10 mM
suggesting that the PG headgroup response is qualitatively captured by simulations, in contrast
to PS in our previous work \cite{antila19}.
and more modest decrease with larger concentrations (Fig. \ref{changesWITHCaClPG}) \cite{borle85}.
This behaviour is similar to that of $\beta$-carbon order parameters of POPC,
but essentially different than observed for POPS, where $\beta$-carbon order parameters increases with
addition of calcium \cite{antila19}. Experimentally measured changes of PG $\alpha$-carbon order parameters upon addition of
calcium are not available.
%upon addition of calcium in experiments,
%exhibiting 
%The behaviour is qualitatively different than in the $\beta$-carbon of PS lipids which increase upon addition of CaCl$_2$.
Lipid17 and Slipids force fields correctly capture the PG $\beta$-carbon order parameter response to CaCl$_2$
even thought the binding affinity was too large based on the comparison of PC headgroup order parameter changes with experiments.
While applying ECC to Lipid17 improved the PC headgroup order parameter response and binding affinity,
the response of PG $\beta$-carbon order parameter to calcium is too weak in this model.
The response of PG $\alpha$-carbon order parameters to CaCl$_2$ differs between force fields,
but experimental data to evaluate these predictions is not available. 
\todo{To be finished once we have the new CHARMM simulations and conformational changes of PG analyzed.}
% The result is similar to the $\sim$200~ns simulations with PC lipids in previous work \cite{catte16}.
%However, when simulation was continued for $\mu$s, the binding affinity substantially increased
%and interpretation was that calcium overbinds to PC lipid in CHARMM36. Therefore, the
%conclusion seems to be similar here, although the new NBfix parameters may complicate
%the situation \todo{The status of NBfix parameters in these simulations should be checked.}.
